Burning sage: Reversing the curse of dimensionality in the visualisation of high-dimensional data

### 2020-09-23 paper/paper.Rmd
gene expression

Four applications of the sage tour are shown here, which illustrate the benefit of using the reverse transformation for examining data using a tour: single cell RNA-seq, classifying hand-sketched images, comparing physics experiments, and the classical pollen data which is used to illustrate the effect of parameter choices.

@@

In the analysis of single cell RNA-seq data, cluster analysis is performed to
detect cell types and find expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis) [@Amezquita2020-at]. Generally, for cluster verification, analysts use embedding methods
like t-SNE to verify the placement and meaning of clusters from a clustering algorithm.

Here we compare the sage display and regular tour display on mouse retinal single cell RNA-seq data from
@Macosko2015-ot. The raw data consists of a 49,300  cells and was downloaded
using the using the `scRNAseq` Bioconductor package [@scRNAseq-d]. We
use a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed quality control using the `scater` package [@McCarthy2017]
and we log-transformed and normalised the expression 
values and selected highly variable genes (HVGs) using `scran` 
[@Lun2016]. The top ten percent of most HVGs were used as features to subset
the normalised expression matrix and compute principal components.
Using the first 25 PCs we built a shared nearest neighbours graph (with $k = 10$) and 
clustered this graph using Louvain clustering, resulting in 11 clusters being formed [@Blondel2008-bx].

We tour the first five PCs (approximately 20% of the variance in expression),
on a weighted subsample of cells based on their cluster membership - 4,590 cells. When running the sage display, we set $\gamma = 3$, fixing the effective dimensionality of the data to $p_{\mathrm{eff}} = 15$. For the displays we have rescaled the PCs so they each have zero mean and unit variance. Here we are interested in mitigating the effects of overplotting by focusing on three clusters that were identified by graph based clustering but appear to overlap on PC plots.

@@

Figure \ref{fig:mouse-cluster} shows selected frames from the animations available in the supplementary material. The still views serve to illustrate the main points, but we encourage the reader to look at the tour animations to fully appreciate the advantage of the sage display.

Using the default tour display (Figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation. The points in the dark green cluster are overlapping those found in the yellow and blue clusters; and it is difficult to see if there is any separation between the blue and yellow clusters.

On the other hand using the sage display (Figure \ref{fig:mouse-cluster}, bottom), expands the center of projection, and results in the
three clusters appearing more separable than the default display. The dark green cluster becomes more visible; the outlier cell that is far away from the rest of the clustered points is more apparent. Moreover, the relative positions of the yellow and blue clusters centers are easier to see, those clusters are distinct from the dark green cluster, yet in most frames they appear to overlap and jumble together, indicating that perhaps those clusters should be merged together.

>>>>

To illustrate the benefit of using the reverse transformation for examining data using a tour, four applications are shown: clustering of single cell RNA-seq, classifying hand-sketched images, comparing physics experiments, and the classical pollen data. The pollen data example is used to illustrate the effect of parameter choices in the sage tour.

@@

In the analysis of single cell RNA-seq data, cluster analysis is performed to
detect cell types and characterise the expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis) [@Amezquita2020-at]. Generally, for cluster verification, analysts use embedding methods
like t-SNE to verify the placement and meaning of clusters from a clustering algorithm. An alternative is to use a tour on a small number of principal components to examine the clusters relative to gene expression.

Here we compare the sage display and regular tour display on mouse retinal single cell RNA-seq data from
@Macosko2015-ot. The raw data consists of a 49,300  cells and was downloaded
using the using the `scRNAseq` Bioconductor package [@scRNAseq-d]. We
use a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): quality control was perfirmed using the `scater` package [@McCarthy2017]
and `scran` 
[@Lun2016] was used to transform and normalise the expression 
values and select highly variable genes (HVGs). The top ten percent of the most HVGs were used as features to subset
the normalised expression matrix and compute the principal components.
Using the first 25 PCs we built a shared nearest neighbours graph (with $k = 10$) and 
clustered this graph using Louvain clustering, resulting in 11 clusters being formed [@Blondel2008-bx].

A tour is run on the first five PCs (approximately 20% of the variance in expression),
on a weighted subsample of cells based on their cluster membership - 4,590 cells. For the sage display, we set $\gamma = 3$, fixing the effective dimensionality of the data to $p_{\mathrm{eff}} = 15$. The PCs are scaled to have zero mean and unit variance. Here we focus on comparing three of the clusters. In the PC plots they look very similar, begging the question whether they should be considered to be separate groups. Figure \ref{fig:mouse-cluster} shows selected frames a default tour (top row) and the sage tour (bottom row). The columns show the same projection, with the difference being the the sage transformation is applied in the sage tour projections. The full animations are available in the supplementary material. The static plots serve to illustrate the main points, but we encourage the reader to look at the tour animations to fully appreciate the advantage of the sage display.

@@

Using the default tour display (Figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation. The points in the dark green cluster are overlapping those found in the yellow and blue clusters; and it is difficult to see if there is any separation between the blue and yellow clusters. In contrast, the sage display (Figure \ref{fig:mouse-cluster}, bottom), expands the center of projection, and results in the differences between the
three clusters being more visible. <!--The dark green cluster becomes more visible; the outlier cell that is far away from the rest of the clustered points is more apparent.--> Particularly, the relative positions of the yellow and blue clusters are easier to see. While these clusters are distinct from the dark green cluster, in most frames they are still overlapping and mixed together, providing evidence that it may be appropriate to consider them a single cluster. Conversely, it can be seen that the dark green cluster is distinctly separated from the other two in some projections.  The sage tour makes these comparisons a little easier.

### 2020-09-23 paper/paper.tex
gene expression

Figure \ref{fig:cdf} compares these two quantities (equations
\ref{eq:cdf} and \ref{eq:vp}), for \(p=3, 10, 100\). At left is
\(v_{2} (r; p, R)\) and on the right is \(v_{p} (r; p, R)\). The
function shapes change in opposite directions as \(p\) increases:
\(v_{2} (r; p, R)\) peaks earlier and \(v_{p} (r; p, R)\) gets flatter.
This is the paradox of the curse of dimensionality in that projected
volume at the center increases with \(p\).

@@

The full radial transformation is therefore given by \begin{equation}
r'_y = v_2^{-1} (v_2(r_y; p, R); 2, R) =  R \sqrt{v_2(r_y; p, R)} = R \sqrt{1-\left(1-\left(\frac{r_y}{R}\right)^2\right)^{p/2}}.
\label{eq:resc}
\end{equation}

The relation between \(r'_y\) and \(r_y\) depends on the number of
dimensions \(p\), and is illustrated for selected values in Figure
\ref{fig:radii}. We see that the transformation is approximately linear
near the center. As \(p\) increases it becomes non-linear faster, and
e.g.~for \(p=10\) the points with radius \(r_y>0.5\) will already be
highly distorted and pushed out towards the last eighth in \(r'_y\).
Figure \ref{fig:circles} demonstrates this for different values of \(p\)
by showing equidistant circles for which the radius has been transformed
according to Eq. \ref{eq:resc}.

@@

The transformation in Eq. \ref{eq:resc} is fixed for a given input
dataset, by evaluating the number of dimensions \(p\) and the maximum
distance from the center \(R\). However, in realistic applications we
may wish to trim the projected data or tune the transformation. In
practice, a combination of both adjustments can be used to further zoom
in on the center of the distribution, or alternatively, to soften the
transformation.

@@

We allow tuning \(p_{\mathrm{eff}} = \gamma p\) by selecting the scaling
parameter \(\gamma\). By default, \(\gamma=1\) and
\(p_{\mathrm{eff}}=p\). When \(\gamma<1\) the rescaling will be softer,
and \(\gamma>1\) results in more aggressive rescaling than suggested by
\(p\) alone. Note that when \(p_{\mathrm{eff}} < 2\) we actually invert
the behavior and shift the focus away from the center, in general this
is not wanted.

@@

While the radial transformation can in general be used with any
low-dimensional display that suffers from data piling, it is most useful
when combined with a dynamic display showing a sequence of interpolated
low-dimensional projections obtained when running a tour. We have
implemented it as a new display method \texttt{display\_sage} in the
\texttt{tourr} package \citep{tourr} in R \citep{rref}.

@@

In the analysis of single cell RNA-seq data, cluster analysis is
performed to detect cell types and find expression of genes that define
those cell types, and the relative orientation of the cell types to each
other (trajectory analysis) \citep{Amezquita2020-at}. Generally, for
cluster verification, analysts use embedding methods like t-SNE to
verify the placement and meaning of clusters from a clustering
algorithm.

Here we compare the sage display and regular tour display on mouse
retinal single cell RNA-seq data from \citet{Macosko2015-ot}. The raw
data consists of a 49,300 cells and was downloaded using the using the
\texttt{scRNAseq} Bioconductor package \citep{scRNAseq-d}. We use a
standard workflow for pre-processing and normalizing this data
(described by \citet{Amezquita2020-at}): we performed quality control
using the \texttt{scater} package \citep{McCarthy2017} and we
log-transformed and normalised the expression values and selected highly
variable genes (HVGs) using \texttt{scran} \citep{Lun2016}. The top ten
percent of most HVGs were used as features to subset the normalised
expression matrix and compute principal components. Using the first 25
PCs we built a shared nearest neighbours graph (with \(k = 10\)) and
clustered this graph using Louvain clustering, resulting in 11 clusters
being formed \citep{Blondel2008-bx}.

We tour the first five PCs (approximately 20\% of the variance in
expression), on a weighted subsample of cells based on their cluster
membership - 4,590 cells. When running the sage display, we set
\(\gamma = 3\), fixing the effective dimensionality of the data to
\(p_{\mathrm{eff}} = 15\). For the displays we have rescaled the PCs so
they each have zero mean and unit variance. Here we are interested in
mitigating the effects of overplotting by focusing on three clusters
that were identified by graph based clustering but appear to overlap on
PC plots.

@@

Figure \ref{fig:mouse-cluster} shows selected frames from the animations
available in the supplementary material. The still views serve to
illustrate the main points, but we encourage the reader to look at the
tour animations to fully appreciate the advantage of the sage display.

Using the default tour display (Figure \ref{fig:mouse-cluster}, top),
the three clusters (dark green, blue, and yellow) are obscured by points
in other clusters as we move through the frames of the animation. The
points in the dark green cluster are overlapping those found in the
yellow and blue clusters; and it is difficult to see if there is any
separation between the blue and yellow clusters.

On the other hand using the sage display (Figure
\ref{fig:mouse-cluster}, bottom), expands the center of projection, and
results in the three clusters appearing more separable than the default
display. The dark green cluster becomes more visible; the outlier cell
that is far away from the rest of the clustered points is more apparent.
Moreover, the relative positions of the yellow and blue clusters centers
are easier to see, those clusters are distinct from the dark green
cluster, yet in most frames they appear to overlap and jumble together,
indicating that perhaps those clusters should be merged together.

>>>>

Figure \ref{fig:cdf} compares these two quantities (Eq. \ref{eq:cdf} and
\ref{eq:vp}), for \(p=3, 10, 100\). At left is \(v_{2} (r; p, R)\) and
on the right is \(v_{p} (r; p, R)\). The function shapes change in
opposite directions as \(p\) increases: \(v_{2} (r; p, R)\) peaks
earlier and \(v_{p} (r; p, R)\) gets flatter. This is the paradox of the
curse of dimensionality in that projected volume at the center increases
with \(p\).

@@

The full radial transformation is therefore given by \begin{equation}
r'_y = v_2^{-1} (v_2(r_y; p, R); 2, R) =  R \sqrt{v_2(r_y; p, R)} = R \sqrt{1-\left(1-\left(\frac{r_y}{R}\right)^2\right)^{p/2}}
\label{eq:resc}
\end{equation}

The relation between \(r'_y\) and \(r_y\) depends on the number of
dimensions \(p\), and is illustrated for selected values in Figure
\ref{fig:radii}. We see that the transformation is approximately linear
near the center. As \(p\) increases it becomes non-linear faster, and
for \(p=10\), for example, the points with radius \(r_y>0.5\) will
already be highly distorted and pushed out towards the last eighth in
\(r'_y\). Figure \ref{fig:circles} demonstrates this for different
values of \(p\) by showing equidistant circles for which the radius has
been transformed according to Eq. \ref{eq:resc}.

@@

The transformation in Eq. \ref{eq:resc} is fixed for a given input
dataset, by evaluating the number of dimensions \(p\) and the maximum
distance from the center \(R\). However, in practice we may wish to trim
the projected data or tune the transformation. A combination of both
adjustments can be used to further zoom in on the center of the
distribution, or alternatively, to soften the transformation.

@@

We allow tuning \(p_{\mathrm{eff}} = \gamma p\) by selecting the scaling
parameter \(\gamma\). By default, \(\gamma=1\) and
\(p_{\mathrm{eff}}=p\). When \(\gamma<1\) the rescaling will be softer,
and \(\gamma>1\) results in more aggressive rescaling than suggested by
\(p\) alone. Note that when \(p_{\mathrm{eff}} < 2\) we actually invert
the behavior and shift the focus away from the center, in general this
is not recommended.

@@

While the radial transformation can in general be used with any
low-dimensional display that suffers from data crowding, it is most
useful when combined with a dynamic display showing a sequence of
interpolated low-dimensional projections obtained when running a tour.
We have implemented it as a new display method called
\texttt{display\_sage} in the \texttt{tourr} package \citep{tourr} in R
\citep{rref}.

@@

In the analysis of single cell RNA-seq data, cluster analysis is
performed to detect cell types and characterise the expression of genes
that define those cell types, and the relative orientation of the cell
types to each other (trajectory analysis) \citep{Amezquita2020-at}.
Generally, for cluster verification, analysts use embedding methods like
t-SNE to verify the placement and meaning of clusters from a clustering
algorithm. An alternative is to use a tour on a small number of
principal components to examine the clusters relative to gene
expression.

Here we compare the sage display and regular tour display on mouse
retinal single cell RNA-seq data from \citet{Macosko2015-ot}. The raw
data consists of a 49,300 cells and was downloaded using the using the
\texttt{scRNAseq} Bioconductor package \citep{scRNAseq-d}. We use a
standard workflow for pre-processing and normalizing this data
(described by \citet{Amezquita2020-at}): quality control was perfirmed
using the \texttt{scater} package \citep{McCarthy2017} and
\texttt{scran} \citep{Lun2016} was used to transform and normalise the
expression values and select highly variable genes (HVGs). The top ten
percent of the most HVGs were used as features to subset the normalised
expression matrix and compute the principal components. Using the first
25 PCs we built a shared nearest neighbours graph (with \(k = 10\)) and
clustered this graph using Louvain clustering, resulting in 11 clusters
being formed \citep{Blondel2008-bx}.

A tour is run on the first five PCs (approximately 20\% of the variance
in expression), on a weighted subsample of cells based on their cluster
membership - 4,590 cells. For the sage display, we set \(\gamma = 3\),
fixing the effective dimensionality of the data to
\(p_{\mathrm{eff}} = 15\). The PCs are scaled to have zero mean and unit
variance. Here we focus on comparing three of the clusters. In the PC
plots they look very similar, begging the question whether they should
be considered to be separate groups. Figure \ref{fig:mouse-cluster}
shows selected frames a default tour (top row) and the sage tour (bottom
row). The columns show the same projection, with the difference being
the the sage transformation is applied in the sage tour projections. The
full animations are available in the supplementary material. The static
plots serve to illustrate the main points, but we encourage the reader
to look at the tour animations to fully appreciate the advantage of the
sage display.

@@

Using the default tour display (Figure \ref{fig:mouse-cluster}, top),
the three clusters (dark green, blue, and yellow) are obscured by points
in other clusters as we move through the frames of the animation. The
points in the dark green cluster are overlapping those found in the
yellow and blue clusters; and it is difficult to see if there is any
separation between the blue and yellow clusters. In contrast, the sage
display (Figure \ref{fig:mouse-cluster}, bottom), expands the center of
projection, and results in the differences between the three clusters
being more visible. Particularly, the relative positions of the yellow
and blue clusters are easier to see. While these clusters are distinct
from the dark green cluster, in most frames they are still overlapping
and mixed together, providing evidence that it may be appropriate to
consider them a single cluster. Conversely, it can be seen that the dark
green cluster is distinctly separated from the other two in some
projections. The sage tour makes these comparisons a little easier.
